---
title: "CA-mlr"
author: "jeysenbach"
date: "7/30/2020"
output: html_document
---
```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tswge)
library(ggthemes)
library(nnfor)
library(zoo)
library(vars)
```

```{r, message=FALSE}
#Import Data obtained from ca.gov
CA <- read.csv("https://raw.githubusercontent.com/JaclynCoate/6373_Time_Series/master/TermProject/Data/CA_COVID_7.16.20.csv", header = T)
#Re-format Date
CA$date <- as.Date(CA$date, format = "%m/%d/%Y")
head(CA)

Totalhosp <- (CA$hospitalized_covid_confirmed_patients + CA$hospitalized_suspected_covid_patients)
```

Since we are predicting Hospitalizations, which we determined has no notable seasonality in the CA data, the number of new cases could be transformed into a more useful predictor by calculating the average of the past 7 days.

```{r}

newcases_7dayavg <- zoo::rollmean(CA$newcountconfirmed, k=7, align = "right")
CA$newcases_7dayavg <- c(CA$newcountconfirmed[1:6],newcases_7dayavg)

#Replot - Avg of last 7 days New Confirmed Cases
ggplot(data=CA, aes(x=date, y=CA$newcases_7dayavg, group=1)) + 
  geom_line(color="gold2") + ggtitle("Last 7 Day Avg Confirmed COVID-19 Cases in CA") + 
  scale_x_date(date_labels = "%b") + xlab("") + ylab("Count") +           theme_fivethirtyeight()
```

The MLR with correlated errors model is calculated using the smoothed version of the new cases data. We first need to forecast new cases, so that we can include them in our prediction of the hospitalizations forecast.

```{r}
#Cut out the zeros at the beginning of Totalhosp and create equal length variable for new cases that lines up with hospital data
Totalhosp1 <- Totalhosp[12:119]
newcases_7dayavg1 <- CA$newcases_7dayavg[5:112]
```

I used this mlp forecast of new cases for the forecast of hospitalizations

```{r}
newcases_7dayavg2 <- newcases_7dayavg[83:108]

newcases_7dayavg2_ts <- ts(newcases_7dayavg2) #create time series object
fit_mlp_cases = mlp(newcases_7dayavg2_ts, reps = 50, comb = "mean") #mlp fit
fit_mlp_cases
fore_mlp_cases <- forecast(fit_mlp_cases, h=7) #univariate 1-week mlp forecast
plot(fore_mlp_cases)
```


The next step is to generate the MLR model that predicts hospitalizations.

### MODEL EXCLUDING TIME

```{r}
#Check for lag between hospitalizations and new cases
ccf(newcases_7dayavg1, Totalhosp1) #no lagging needed based on ccf plot

#fit linear model for predicting hospitalization
mlr_fit <- lm(Totalhosp1~newcases_7dayavg1)
plot(mlr_fit$residuals)
#fit residuals
aic5.wge(mlr_fit$residuals) 
#low p/q models as expected. The top pick of ARMA(1,1) should be reasonable
fit1 = arima(Totalhosp1, order=c(1,0,1), xreg=newcases_7dayavg1)
fit1 #aic = 1436

#check residuals of model
plot(fit1$residuals)
acf(fit1$residuals) #appears to have no autocorrelation
lj24 <- ljung.wge(fit1$residuals, p=1, q=1)
lj24$pval
lj48 <- ljung.wge(fit1$residuals, p=1, q=1, K=48)
lj48$pval
#Ljung-Box test fails to reject H0 - no evidence against white noise

#forecast hospitalizations using prior forecast of 7 day avg new cases by mlp model
next7 = data.frame(new_cases_avg = fore_mlp_cases$mean)
f_mlr <- predict(fit1, newxreg = next7, n.ahead = 7)

plot(seq(1,108,1), Totalhosp1, type = "l",xlim = c(0,115),ylim=c(4000,9000), xlab="days", ylab = "COVID-Related Hospitalized Patients", main = "7 Day Forecast - Linear Regression with Corr Errors Model")
lines(seq(109,115,1), f_mlr$pred, type = "l", col = "red")
```



### MODEL INCLUDING TIME

```{r}
#fit linear model for predicting hospitalization, including time as a variable
Time <- seq(1,108,1)
tmlr_fit <- lm(Totalhosp1~newcases_7dayavg1+Time)
plot(tmlr_fit$residuals)
#fit residuals
aic5.wge(tmlr_fit$residuals) 
#The top pick is ARMA(1,1) again
fit1t = arima(Totalhosp1, order=c(1,0,1), xreg=cbind(newcases_7dayavg1,Time))
fit1t #aic =1433, slightly better with time included

#check residuals of model
plot(fit1t$residuals)
acf(fit1t$residuals) #appears to have no autocorrelation
lj24 <- ljung.wge(fit1t$residuals, p=1, q=1)
lj24$pval
lj48 <- ljung.wge(fit1t$residuals, p=1, q=1, K=48)
lj48$pval
#Ljung-Box test fails to reject H0 - no evidence against white noise

#forecast hospitalizations using prior forecast of 7 day avg new cases by mlp model
next7 = data.frame(new_cases_avg = fore_mlp_cases$mean, Time = seq(109,115,1))
f_mlr <- predict(fit1t, newxreg = next7, n.ahead = 7)

plot(seq(1,108,1), Totalhosp1, type = "l",xlim = c(0,115),ylim=c(4000,9000), xlab="days", ylab = "COVID-Related Hospitalized Patients", main = "7 Day Forecast - Linear Regression with Corr Errors Model")
lines(seq(109,115,1), f_mlr$pred, type = "l", col = "red")
```


The model with time included as a variable in the MLR had a slightly lower aic, but the forecast is really anyone's guess. For a seven-day prediction, the second (with time included in linear model variables) looks like a reasonable extention of the current trend.
