---
title: "CA-mlr"
author: "jeysenbach"
date: "7/30/2020"
output: html_document
---
```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tswge)
library(ggthemes)
library(nnfor)
library(zoo)
library(vars)
```

```{r, message=FALSE}
#Import Data obtained from ca.gov
CA <- read.csv("https://raw.githubusercontent.com/JaclynCoate/6373_Time_Series/master/TermProject/Data/CA_COVID_7.16.20.csv", header = T)
#Re-format Date
CA$date <- as.Date(CA$date, format = "%m/%d/%Y")
head(CA)

Totalhosp <- (CA$hospitalized_covid_confirmed_patients + CA$hospitalized_suspected_covid_patients)
```


# OLD STUFF - needed to create correct variables...

Since we are predicting Hospitalizations, which we determined has no notable seasonality in the CA data, the number of new cases could be transformed into a more useful predictor by calculating the average of the past 7 days.

```{r}

newcases_7dayavg <- zoo::rollmean(CA$newcountconfirmed, k=7, align = "right")
CA$newcases_7dayavg <- c(CA$newcountconfirmed[1:6],newcases_7dayavg)

#Replot - Avg of last 7 days New Confirmed Cases
ggplot(data=CA, aes(x=date, y=CA$newcases_7dayavg, group=1)) + 
  geom_line(color="gold2") + ggtitle("Last 7 Day Avg Confirmed COVID-19 Cases in CA") + 
  scale_x_date(date_labels = "%b") + xlab("") + ylab("Count") +           theme_fivethirtyeight()
```

```{r}
#Cut out the zeros at the beginning of Totalhosp and create equal length variable for new cases that lines up with hospital data
Totalhosp1 <- Totalhosp[12:119]
newcases_7dayavg1 <- CA$newcases_7dayavg[5:112]
```

```{r}
newcases_7dayavg2 <- newcases_7dayavg[83:108]
```


### California: Vector AR Models

We can use the same variables to model using VAR.

```{r}
#Create matrix of variables
var_matrix1 <- cbind(newcases_7dayavg1, Totalhosp1)

#VAR modeling
#We will model as if the data is stationary
VARselect(var_matrix1, lag.max = 10, type = "both") #AIC picks 9, BIC picks 1
vfit1_1 <- VAR(var_matrix1,p=9,type = "both")

#7 Day forecast
vpreds1_7 <- predict(vfit1_1,n.ahead = 7)
vpreds1_7$fcst$Totalhosp1

#Plot 7 day forecast
Time <- seq(1,108,1)
plot(Time, Totalhosp1, type = "l",xlim = c(0,115),  ylim = c(4000,10000), ylab = "COVID-Related Hospitalized Patients", main = "7 Day Forecast - VAR Model")
lines(seq(109,115,1), vpreds1_7$fcst$Totalhosp1[,1], type = "l", col = "red")
```

Again, we'll consider a 90 day forecast based on this model to be unrealisitc based on our expectations.

```{r}
#90 Day forecast
vpreds1_90 <- predict(vfit1_1,n.ahead = 90)

#Plot 7 day forecast
plot(Time, Totalhosp1, type = "l",xlim = c(0,198),  ylim = c(4000,20000), ylab = "COVID-Related Hospitalized Patients", main = "90 Day Forecast - VAR Model")
lines(seq(109,198,1), vpreds1_90$fcst$Totalhosp1[,1], type = "l", col = "red")
```
