---
title: "US Data EDA"
author: "Jaclyn Coate & Josh Eysenbach"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(skimr)
```

# US COVID Data

## EDA 

### US Covid Data
We leveraged the [The Covid Tracking Project](https://covidtracking.com/) for analyzing the COVID impact on the United States. The COVID Tracking Project is being utilized by many leading and trusted groups such as  the CDC, HHS, and Johns Hopkins.

Using the simple read.csv command we load the data from our [Data](https://github.com/JaclynCoate/6373_Time_Series/tree/master/TermProject/Data) folder for our [Term Project](https://github.com/JaclynCoate/6373_Time_Series/tree/master/TermProject) Github repository.

```{r load data}
us <- read.csv("https://raw.githubusercontent.com/JaclynCoate/6373_Time_Series/master/TermProject/Data/USdaily7.19.csv", header = T, strip.white = T)
str(us)
```

### US: Available Variables
1. *date* - Date on which data was compiled by The COVID Tracking Project.
2. *states* - Number of states and territories included in the US dataset for this day.
3. *positive* - Total number of people with confirmed OR probable COVID-19 reported by the state or territory (per the expanded CSTE case definition of April 5th, 2020 approved by the CDC). A confirmed case is a person who has a positive test result from an FDA approved diagnostic molecular test. A probable case is a person who either has presentable symptoms WITH epidemiological evidence or has BOTH a positive presumptive laboratory test AND also EITHER presentable symptoms OR epidemiological evidence. Epidemiological evidence refers either to close-proximity contact with a known case or travel history to an area with high disease incidence. According to the guidelines, FDA approved antibody and antigen tests are considered part of presumptive laboratory evidence.
4. *negative* - Individuals with a completed viral test that returned a negative result. For states / territories that do not report this number directly, we compute it using one of several methods, depending on which data points the state provides.
5. *pending* - Total number of viral tests that have not been completed as reported by the state or territory.
6. *hospitalizedCurrently* - Individuals who are currently hospitalized with COVID-19. Definitions vary by state / territory. Where possible, we report hospitalizations with confirmed or probable COVID-19 cases per the expanded CSTE case definition of April 5th, 2020 approved by the CDC.
7. *hospitalizedCumulative* - Total number of individuals who have ever been hospitalized with COVID-19. Definitions vary by state / territory. Where possible, we report hospitalizations with confirmed or probable COVID-19 cases per the expanded CSTE case definition of April 5th, 2020 approved by the CDC.
8. *inIcuCurrently* - Individuals who are currently hospitalized in the Intensive Care Unit with COVID-19. Definitions vary by state / territory. Where possible, we report patients in the ICU with confirmed or probable COVID-19 cases per the expanded CSTE case definition of April 5th, 2020 approved by the CDC.
9. *inIcuCumulative* - Total number of individuals who have ever been hospitalized in the Intensive Care Unit with COVID-19. Definitions vary by state / territory. Where possible, we report patients in the ICU with confirmed or probable COVID-19 cases per the expanded CSTE case definition of April 5th, 2020 approved by the CDC.
10. *onVentilatorCurrently* - Individuals who are currently hospitalized under advanced ventilation with COVID-19. Definitions vary by state / territory. Where possible, we report patients on ventilation with confirmed or probable COVID-19 cases per the expanded CSTE case definition of April 5th, 2020 approved by the CDC.
11. *onVentilatorCumulative* - Total number of individuals who have ever been hospitalized under advanced ventilation with COVID-19. Definitions vary by state / territory. Where possible, we report patients on ventilation with confirmed or probable COVID-19 cases per the expanded CSTE case definition of April 5th, 2020 approved by the CDC.
12. *recovered* - Total number of people that are identified as recovered from COVID-19. States provide disparate definitions on what constitutes as a “recovered” COVID-19 case. Types of “recovered” cases include those who are discharged from hospitals, released from isolation after meeting CDC guidance on symptoms cessation, or those who have not been identified as fatalities after a number of days (30 or more) post disease onset. Specifics vary for each state or territory.
13. *dateChecked* - Deprecated. This is an old label for *lastUpdateEt*.
14. *death* - Total fatalities with confirmed OR probable COVID-19 case diagnosis (per the expanded CSTE case definition of April 5th, 2020 approved by the CDC). In states where the information is available, it only tracks fatalities with confirmed OR probable COVID-19 case diagnosis where COVID-19 is an underlying cause of death according to the death certificate based on WHO guidelines.
15. *hospitalized* - Deprecated. Old label for hospitalizedCumulative.
16. *lastModified* - Deprecated. Old label for lastUpdateET.
17. *total* - Deprecated. Computed by adding *positive*, *negative*, and *pending* values.
18. *totalTestRestuls* - Currently computed by adding positive and negative values to work around reporting lags between positives and total tests and because some states do not report totals. Deprecated in the API and soon to be replaced on the website as well.
19. *posNeg* - Deprecated. Computed by adding *positive* and *negative* values.
20. *deathIncrease* - Increase in *death* computed by subtracting the value of *death* for the previous day from the value of *death* for the current day.
21. *hospitalizedIncrease* - Increase in *hospitalizedCumulative* computed by subtracting the value of *hospitalizedCumulative* for the previous day from the value of *hospitalizedCumulative* for the current day.
22. *negativeIncrease* - Increase in *negative* computed by subtracting the value of *negative* for the previous day from the value for *negative* from the current day.
23. *positiveIncrease* - Increase in *positive* computed by subtracting the value of *positive* from the previous day from the value of *positive* for the current day.
24. *totalTestRestulsIncrease* - Deprecated. Increase in *totalTestResults* computed by subtracting the value of *totalTestResults* for the previous day from the value of *totalTestResults* for the current day.
25. *hash* - A hash for this record

### Variable Evaluation and Formatting

```{r}
us <- transform(us, date = as.Date(as.character(date), "%Y%m%d"))
head(us)
```

### Logical Variable Removal
During this project we will be trying to surface those metrics that properly evaluate the severity of COVID in the US. In order to simplify our data set we will remove all variables that do not directly link to metrics dealing with COVID or variables that are depcitign the same information. E.g. _totalTestResults_ and _postNeg_.

```{r}
us <- subset(us, select = -c(states, dateChecked, hospitalized, lastModified, total, posNeg, totalTestResultsIncrease, hash))
head(us)
```

### NA Evaluation

Upon evaluating the columns that contain not available information, we can see in the early dates of data collection the US had a low count of COVID cases. Since it is likely that those metrics were actually zero since they weren't even being collected at that time, we will repalce all NAs with zeros.

```{r}
us[is.na(us)] <- 0
tail(us)
```

### Zero variance variable check
All variables show a variance therefore we will keep all in the model for now

```{r zero variable check}
skim(us)
```

## COVID Case Cumulative Variable Review
Below we will use a basic lines graph by day to see the general trend of some of the basic cumulative variables that are provided by our raw data set. The cumulative variables are interesting to see the general trend of the data. However, these would not be beneficial from a forecasting iniative due to the fact that these will just continue to trend upwards as the virus spreads. It gives no real insight that could be used for our COVID relief efforts for hospitals. 

### Cumulative COVID Cases in US
We can see that the total cumulative COVID cases are increasing overtime with what appears to be no reprieve.

```{r}
library(ggplot2)
library(ggthemes)

ggplot(data = us, aes(x=date, y=positive))+
  geom_line(color="black")+
  labs(title = "Cumulative COVID Cases US", y = "Millions", x = "") +
  theme_fivethirtyeight()
```

### Cumulative COVID Hospitalized Cases in US
The total hospitilizations due to covid appear to also increase over time. We do see a little more movement in this graph, however, overall clearly increasing.

```{r}
ggplot(data = us, aes(x=date, y=hospitalizedCumulative))+
  geom_line(color="orange")+
  labs(title = "Cumulative COVID Hospitalized Cases US", y = "Millions", x = "") +
  theme_fivethirtyeight()
```

### Cumulative COVID Related Deaths in US
The cumulative deaths are also increasing over time. We can clearly see what appears to be an upward trend.

```{r}
ggplot(data = us, aes(x=date, y=death))+
  geom_line(color="dark red")+
  labs(title = "Cumulative COVID Death Cases US", y = "100 Thousands", x = "") +
  theme_fivethirtyeight()
```

## New and Currently COVID Case Variable Review
In an attempt to look at a more micro view of the COVID data we review some of the new and cumulative variables below. We are looking to surface those variables that will be the most benficial in forecasting the severity of cases and therefore hospital resources moving forward.

### New Confirmed COVID in US
When we start to investigate at a more micro level and review something like the daily new cases we see more trending behavior and something that can be more properly evaluated.
Newly confirmed cases have a little more inforamtion to tell use from a trend perspective. We can see some pseudo-cyclic behavior alongside some heavy wandering behavior.

```{r}
ggplot(data = us, aes(x=date, y=positiveIncrease))+
  geom_line(color="black")+
  labs(title = "New COVID Cases US", y = "Thousands", x = "") +
  theme_fivethirtyeight()
```

### Currently COVID Hospitalizations in US
When measuring the severity of cases, we wanted to make sure that we are including the most severe cases; we believe that a good indicator of this would be a hospitalization metric. This is not only because the severe cases would end up hospitalized, but because of the intial fear of resources and space, it would also be a key forecast for medical communities to be prepared for potential next round of patients.

We have seen that the research suggests 'that most people who contract the new coronavirus develop mild cases of Covid-19' from an article released in June that can be found [here](https://www.advisory.com/daily-briefing/2020/06/01/asymptomatic-patients). Due to this fact of likely a-symptomatic patients, hospitalization metrics reveal themselves to be a more accurate representation of severe cases.

Originally we were going to evaluate the new COVID hospitalization cases. However, upon closer review it became clear that large states that were hit the hardest were failling to reporting this metric. We were able to confirm this analysis when we compared the New versus Current metrics and see that there is a large spike in the hospitalization cases overall, but we are not seeing the same results reflected in the New COVID cases metric. Therefore, we err on the side of caution and move forward with modeling our Currently COVID hospitalized metric. 

Looking at a more micro view of currently hospitalized COVID cases shows us much more behavior that we might be able to evaluate moving forward. Below we can see what appears to be heavy wandering behavior and if we look closely, some additional noise that could be pseudo-cyclic behavior hidden by the large numbers.

```{r}
ggplot(data = us, aes(x=date, y=hospitalizedIncrease))+
  geom_line(color="orange")+
  labs(title = "New COVID Hospitalized Cases US", y = "Thousands", x = "") +
  theme_fivethirtyeight()

ggplot(data = us, aes(x=date, y=hospitalizedCurrently))+
  geom_line(color="orange")+
  labs(title = "Current COVID Hospitalized Cases US", y = "Thousands", x = "") +
  theme_fivethirtyeight()
```

### New COVID Deaths in US
When we review the new death cumulative count we can see what again appears to be a pseudo-cyclic behavior with maybe some slight wandering tendencies.

```{r}
ggplot(data = us, aes(x=date, y=deathIncrease))+
  geom_line(color="dark red")+
  labs(title = "New COVID Death Cases US", y = "Thousands", x = "") +
  theme_fivethirtyeight()
```

## Caculated Variables Creation & Review
In order to review the data at an even smaller level we have created the positive percent metric as well as hospitilication, death, and ICU rates of positive case.

### Positive Percent Rate
When reviewing the daily positivty rate, the metric discussed that is more reported on lately. We see the positivity rate actually decreasing over time. We would expect this outcome since we saw a drastic increase in testing over time as well.

While this metric is better than just new cases to forecast severity (it normalizes to number of tests), it doesn't tell us anything about impact. We tend towards the currently hospitalized metrics because it allows us to take into consideration things such medical resources and why, we as a nation, were attempting to flatten the curve. 

```{r}
us$posRate <- us$positive / us$totalTestResults

ggplot(data = us, aes(x=date, y=posRate))+
  geom_line(color="black")+
  labs(title = "Daily Positivity Rate COVID Testing US", y = "Rate", x = "") +
  theme_fivethirtyeight()
```

### Hospitalization Rate
When reviewing rates we also chose to observe the trend of the hospitalization rate of positively tested COVID patients. We see the hospitalization rate start very low then spike to around 15%. The hospitilization rate seems to hover in the low teens for a few months and then it appears to beging to decrease.

While reviewing this data we determined that this was not a best metric to review moving forward for forecasting severity of cases. The reason being is that individuals who were hospitalized for reasons other than COVID but then returned a positive test would be listed as a COVID patient, even if the virus itself was not affecting them (a-symptomatic.) Therefore we tend towards currently hospitalized as our severity metric, since it takes into account all hospitalizatoin resources and doesn't just try to focus on a metric that appears to be misleading at this time.

```{r}
us$hospRate <- us$hospitalizedCumulative / us$positive

ggplot(data = us, aes(x=date, y=hospRate))+
  geom_line(color="dark orange")+
  labs(title = "Daily Hospitalization Rate COVID Testing US", y = "Rate", x = "") +
  theme_fivethirtyeight()
```

### Recovery Rate
In an opposite pattern of the previous rates we have evaluted we actually see a sharp increase in recovyer rate over time. It appears to be plateauing recently, but what we would expect is for it to continue to increase and provide a final percent of infected individuals who have recovered from COVID.

At first we thought the Recovery Rate would be a great opposite indicator to compare against a severity indicator. However, upon closer investigation we quickly realized that merely a few states were even collecting some of the recovered data. Without an actual number coming from all states this would not be a good refelcting of the recovery rate of COVID over all. This data would not be good to forecast.

```{r}
us$recRate <- us$recovered / us$positive

ggplot(data = us, aes(x=date, y=recRate))+
  geom_line(color="dark green")+
  labs(title = "Daily Recovery Rate COVID Testing US", y = "Rate", x = "") +
  theme_fivethirtyeight()
```

### Death Rate
Like the previous rates we have examined before we see an initial spike in death rate and what appears to be drop off as quickly on the rate of death itself. When we see the death rate hovering right under 4% we should also nte that the flu death rate is about 1%. That puts COVID about 3 times as deadly as the flu.

Since death is the most severe outcome of the COVID virus we have chosen to model this metric as a measure of severity along side currently hospitalized. With these two metrics we feel like we will have a better picture of severity and be able to give our nation a better idea 

```{r}
us$deathRate <- us$death / us$positive

ggplot(data = us, aes(x=date, y=deathRate))+
  geom_line(color="dark red")+
  labs(title = "Daily Death Rate COVID Testing US", y = "Rate", x = "") +
  theme_fivethirtyeight()
```

## Matrix Scatterplot of Surfaced Variables
Trying to create pairs plots for all the variables against eachother would be very difficult to interpret. As suspected there are highly correlated variables since we are graphing things like total postive tests and new positive tests. These variables would be inherently correlated to one another because they are communicating similar data. Since this is the case we will only review a scatterplot of those variables we have surfaced above and will use for forecasting the multivariate testing.

```{r}
library(GGally)
scatter <- subset(us, select = c(positive, negative, hospitalizedCurrently, hospitalizedCumulative, death, totalTestResults, deathIncrease, hospitalizedIncrease, negativeIncrease, positiveIncrease, posRate, deathRate))
ggpairs(scatter)
```
